---
title: "Cluster testing and possible workflow"
author: "Sarah E Taylor"
date: "2024-07-02"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

This document will go through how to use the Umich Great Lakes ARC cluster to run the Locks Of Lineage analysis that runs in RevBayes. The RevBayes software is something that must be loaded and worked with in the terminal.

## Getting into the Cluster in the terminal
ssh taysarah@greatlakes.arc-ts.umich.edu

## 2. Determining Potential Bottlenecks

This step goes through the different parts of the mcmc_MuSSE.Rev script to determine which are the slowest and most computationally intensive to determine what would be best to potentially parallelize.

```{bash Set-up-RevBayes}
#Open RevBayes software in the terminal
~/Desktop/GitHub/LocksofLineage/code/rb

#Run the script in the revbayes program
source("code/file.Rev")
```
These are the basic steps to run RevBayes in the terminal. Each of the RevBayes scripts, ending in .Rev, must have the correct variables defined depending on which trait is being analyzed. Each script has a line (q()) that quits the RevBayes software after the script has run. 

### # Benchmark data reading
```{bash Variables-Read-Data-Benchmarks}
profile("start", "Step 1: Reading_Data")
#Variables
NUM_STATES = 5
DATASET = "ACS"

# Read in the data and the tree
observed_phylogeny <- readTrees("data/RevBayes_Data/pruned_tree.nex")[1]
data <- readCharacterData("data/RevBayes_Data/"+DATASET+"_output.nex")

# Get some useful variables from the data
taxa <- observed_phylogeny.taxa()

profile("stop", "Step 1: Reading_Data")

profile("reportTotalTime", "Step 1: Reading_Data")
```
This is the output for report Total Time:  2.795341 seconds


### Benchmark setting up rates and moves
```{bash Rates-Moves-Benchmark}
profile("start", "Step 2: Rates_Moves")

moves    = VectorMoves()
monitors = VectorMonitors()

# Constant
H = 0.587405

# Create the rates and specify a prior on the diversification and turnover rate
for (i in 1:NUM_STATES) {

     ### Create a loguniform distributed variable for the diversification rate
    speciation[i] ~ dnLoguniform( 1E-6, 1E2)
    speciation[i].setValue( ln(367.0/2.0) / observed_phylogeny.rootAge() )
    moves.append( mvScale(speciation[i],lambda=0.20,tune=true,weight=3.0) )

    ### Create a loguniform distributed variable for the turnover rate
    extinction[i] ~ dnLoguniform( 1E-6, 1E2)
    extinction[i].setValue( speciation[i]/10.0 )
    moves.append( mvScale(extinction[i],lambda=0.20,tune=true,weight=3.0) )

    diversification[i] := speciation[i] - extinction[i]

}

profile("stop", "Step 2: Rates_Moves")

profile("reportTotalTime", "Step 2: Rates_Moves")
```
1.378395 seconds

### Benchmarking the rate matrix

```{bash Rate-Matrix-Benchmark}
profile("start", "Step 3: Rates_Matrix")

# Set up the transition rate matrix for observed states

# Each transition rate between observed states are drawn
# from an exponential distribution with a mean of 10
# character state transitions over the tree.
rate_pr := observed_phylogeny.treeLength() / 10
for ( i in 1:(NUM_STATES*(NUM_STATES-1)) ) {
    transition_rates[i] ~ dnExp(rate_pr)
    moves.append( mvScale(transition_rates[i],lambda=0.20,tune=true,weight=3.0) )
}

# Create the rate matrix for the combined observed and hidden states
rate_matrix := fnFreeBinary( transition_rates, rescaled=false)

profile("stop", "Step 3: Rates_Matrix")

profile("reportTotalTime", "Step 3: Rates_Matrix")
```
 0.891422 seconds

### Benchmarking the set up of the root state frequencies

```{bash}
profile("start", "Step 4: Root_Freqs")

### Create a constant variable with the prior probabilities of each rate category at the root.
rate_category_prior ~ dnDirichlet( rep(1,NUM_STATES) )
moves.append( mvDirichletSimplex(rate_category_prior,tune=true,weight=2) )


### Create the fix parameter for the age of the root set to the observed age
root <- observed_phylogeny.rootAge()


### rho is the probability of sampling species at the present
### fix this to 233/367
rho <- observed_phylogeny.ntips()/367


#### construct a variable for the tree drawn from a birth death process
timetree ~ dnCDBDP( rootAge           = root,
                    speciationRates   = speciation,
                    extinctionRates   = extinction,
                    Q                 = rate_matrix,
                    pi                = rate_category_prior,
                    delta             = 1.0,
                    rho               = rho,
                    condition         = "time")

### clamp the model with the "observed" tree
timetree.clamp( observed_phylogeny )
timetree.clampCharData( data )

profile("stop", "Step 4: Root_Freqs")

profile("reportTotalTime", "Step 4: Root_Freqs")
```

### Benchmarking the model set up and creation of monitors

```{bash}
profile("start", "Step 5: Model_Monitor")

### workspace model wrapper ###
mymodel = model(rate_matrix)

### set up the monitors that will output parameter values to file and screen
monitors.append( mnModel(filename="output/RevBayes/log/SSE_Models/primates_MuSSE_"+DATASET+".log", printgen=1) )
monitors.append( mnJointConditionalAncestralState(tree=timetree, cdbdp=timetree, type="Standard", printgen=1, withTips=true, withStartStates=false, filename="output/RevBayes/log/SSE_Models/primates_MuSSE_"+DATASET+"_anc_states.log") )
monitors.append( mnStochasticCharacterMap(cdbdp=timetree, filename="output/RevBayes/log/SSE_Models/primates_MuSSE_"+DATASET+"_stoch_map.log", printgen=1) )
monitors.append( mnScreen(printgen=10, speciation, extinction) )

profile("stop", "Step 5: Model_Monitor")

profile("reportTotalTime", "Step 5: Model_Monitor")
```

### Benchmarking the model analysis on one generation

```{bash}
profile("start", "Step 6: MCMC_1gen")

# workspace mcmc
mymcmc = mcmc(mymodel, monitors, moves, nruns=2, moveschedule="random", combine="mixed")

# run the MCMC
mymcmc.run(generations=1)

profile("stop", "Step 6: MCMC_1gen")

profile("reportTotalTime", "Step 6: MCMC_1gen")
```


```{r All-color-states-musse-plot}
library(RevGadgets)
library(ggplot2)

# read in and process the log file
musse_file <- paste0("output/RevBayes/log/SSE_Models/primates_MuSSE_ACS.log")
pdata <- processSSE(musse_file)

# plot the rates
plot <- plotMuSSE(pdata) +
        theme(legend.position = c(0.875,0.915),
              legend.key.size = unit(0.4, 'cm'), #change legend key size
              legend.title = element_text(size=8), #change legend title font size
              legend.text = element_text(size=6))

ggsave(paste0("output/RevBayes/plots/BiSSE_div_rates_activity_period.png"),plot, width=5, height=5)
```

```{r}


```

